Geometric variational problems

### 2020-11-13 NSF2020/Description/specificaim3_ver2.tex
small edits

\subsection{Specific Aim 3: Efficient high-order numerical methods for
large-scale simulations}
\label{subsec:specific_aim_3}
% ----------------------------------------------------------------------
The HAP model requires solving exterior Dirichlet problems for the
screened Laplace equation and the Stokes mobility problem at each time
step in complex domains such as in Figure~\ref{fig:domain}. If
discretizing these equations with stencil-based numerical methods, the
computational domain must be truncated, the fluid volume must be
discretized, and artificial boundary conditions must be imposed. The
artificial boundary conditions introduces additional error, and
discretizing the volume results in an excessively large linear system.
Moreover, it is difficult to obtain high-order discretizations when the
boundary surfaces are irregular. Alternatively, since we are solving
constant-coefficient elliptic PDEs, the solution of the PDE can be
represented as layer potentials
\begin{align}
  \label{eq:LP}
  u(\mathbf{x}) = \int_{\partial\Omega} K(\mathbf{x},\mathbf{y})
  \sigma(\mathbf{y})\,dS,
\end{align}
where $\sigma$ is an unknown density function, and
$K(\mathbf{x},\mathbf{y})$ is a derivative of the fundamental solution
of the underlying PDE. By matching the boundary condition at
$\mathbf{x}_0 \in \partial\Omega$ with the limit of
equation~\eqref{eq:LP} as $\mathbf{x}\rightarrow \mathbf{x}_0$, a BIE
for the density function is formed. Numerical methods to solve BIEs have
several advantages over their PDE-based counterpart: by construction,
the layer potential~\eqref{eq:LP} satisfies far-field conditions; only
$\partial\Omega$ needs to be discretized; carefully chosen kernels
$K(\mathbf{x},\mathbf{y})$ result in a well-conditioned linear system
that is solved with a mesh-independent number of GMRES iterations; and
high-order or spectral accuracy is attainable with appropriate
quadrature methods. In our previous work~\cite{Fu2018_SIAM} we
demonstrate that BIEs are a powerful tool to simulate two-dimensional
suspensions of amphiphilic particles. In addition to improving
two-dimensional simulations, we will extend the results to three
dimensions using a standard three-dimensional BIE formulation of the
screened Laplace equation~\cite{ying_2006} and a well-conditioned BIE
formulation of the mobility problem~\cite{manasthesis, rac-gre2016}.

\begin{wrapfigure}[21]{r}{2.5in}
%\vspace*{+15pt}
\centerline{\includegraphics[width=2.5in]{figures/integrands}}
\vspace*{-13pt}
  \caption{{\footnotesize The Fourier modes of
  $K(\mathbf{x_0},\mathbf{y})$ and $K_2(\mathbf{x_0},\mathbf{y})$ where
  $|\mathbf{x_0}| = 0.99$ and $\mathbf{y} \in \partial\Omega$, and
  $\Omega$ is the unit circle. The trapezoid rule applied to the red
  integrand has large amounts of error, but the trapezoid rule is much
  more accurate when applied to the blue integrand. Since the difference
  between the red and blue curves can be accurately computed with the
  barycentric quadrature rule, the quadrature error applied to layer
  potentials for the screen Laplace equation will be uniformly bounded
  in $\Omega$.}}
\label{fig:integrands}
\end{wrapfigure}
Significant challenges to solve BIEs include solving dense linear
systems, developing preconditioning strategies, and developing
quadrature methods for integrands that are nearly singular. These
challenges are described in section~\ref{subsec:NumericalIssues}.
However, numerical errors can still lead to unphysical contact between
bodies. We propose two algorithms to avoid contact in
Section~\ref{subsec:timeStepping}. Then, proposed work for
well-conditioned BIE formulations of fluctuating hydrodynamics are
described in Section~\ref{subsec:fluctuating}.

@@

Discretizations of carefully chosen BIEs can be solved with a
mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}. Therefore, the required CPU
time is proportional to the cost of a matrix-vector multiplication that
can be done in optimal or near-optimal time with the fast multipole
method (FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3,
fmm4, fmm6, fmm7, fmm8}. PI BQ is experienced with applying FMMs to
BIE formulations of the Stokes equation~\cite{qua-bir2014,
bys-sha-qua2020} and the screened Laplace equation~\cite{kro-qua2011,
qua2011}. Alternatively, fast direct solvers can be used to avoid
iteration~\cite{fds1, fds2, fds3, fds4, fds5, fds6, fds7, fds8,
ho2016cpam2, ho2016cpam1, minden2016, minden2017siammms}, but direct
solvers often require a large amount of initial overhead, and this makes
them less practical for problems with moving geometries. Alternatively,
techniques in fast direct solvers can be used to develop efficient
preconditioning strategies such as the inverse Fast Multipole Method
(IFMM)~\cite{cou-pou-dar2017}. PI BQ used the IFMM to precondition
Stokes equations in a porous media~\cite{qua-cou-dar2018}, and a suite
of other preconditioners including sparse approximate
inverses~\cite{che2000} and multigrid~\cite{hem-sch1981, sch1982} will
be investigated.

Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and this results in nearly-singular integrands. Quadrature
methods for these integrands has received a lot of attention in two and
three dimensions~\cite{alpert, kapur, sidi, duffy, bruno1, bruno2,
davis_1984, graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003,
khayat_2005, schwab_1992, ying_2006, beale1, beale2, goodman_1990,
haroldson_1998, lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3,
helsing_2008a, helsing_integral_2009, helsing_tutorial_2012,
klockner2013jcp, qbx2, wala2019jcp, af2018sisc, siegel2018jcp,
rachh2017jcp, ding2019arxiv, bar2014}. The trapezoid rule is the
workhorse for two-dimensional BIEs since it achieves spectral accuracy
when integrands are not nearly-singular~\cite{tre-wei2014}. To address
nearly-singular integrands of two-dimensional BIEs, we will use a {\em
barycentric quadrature rule} that only requires a slight modification of
the trapezoid rule~\cite{ioa-pap-per1991}. This method requires the
layer potential satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs will extend
this quadrature method to layer potentials for the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as
$K(\mathbf{x},\mathbf{y}) = K_1(\mathbf{x},\mathbf{y}) +
K_2(\mathbf{x},\mathbf{y})$, where $K_1(\mathbf{x},\mathbf{y}) =
-\log|\mathbf{x} - \mathbf{y}|$ and $K_2(\mathbf{x},\mathbf{y}) =
K(\mathbf{x},\mathbf{y}) + \log|\mathbf{x} - \mathbf{y}|$. Using this
decomposition, layer potentials involving $K_1(\mathbf{x},\mathbf{y})$
can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and layer potentials involving
$K_2(\mathbf{x},\mathbf{y})$ can be accurately computed with the
trapezoid rule since the kernel is bounded for all $\mathbf{x}$ and
$\mathbf{y}$. The proposed method is demonstrated in
Figure~\ref{fig:integrands} where the amplitude of the Fourier modes of
$K(\mathbf{x_0},\mathbf{y})$ and $K_2(\mathbf{x_0},\mathbf{y})$ are
plotted. The quadrature of the trapezoid rule is the sum of the Fourier
amplitudes not captured at the resolution.

>>>>

\subsection{Specific Aim 3: Efficient high-order numerical methods for
large-scale simulations}
\label{subsec:specific_aim_3}
% ----------------------------------------------------------------------
The HAP model requires solving exterior Dirichlet problems for the
screened Laplace equation and the Stokes mobility problem at each time
step in complex domains such as in Figure~\ref{fig:domain}. If
discretizing these equations with stencil-based numerical methods, the
computational domain must be truncated, the fluid volume must be
discretized, and artificial boundary conditions must be imposed. The
artificial boundary conditions introduce additional error, and
discretizing the volume results in an excessively large linear system.
Moreover, it is difficult to obtain high-order discretizations when the
boundary surfaces are irregular. Alternatively, since we are solving
constant-coefficient elliptic PDEs, the solution of the PDEs can be
represented as layer potentials
\begin{align}
  \label{eq:LP}
  u(\mathbf{x}) = \int_{\partial\Omega} K(\mathbf{x},\mathbf{y})
  \sigma(\mathbf{y})\,dS,
\end{align}
where $\sigma$ is an unknown density function, and
$K(\mathbf{x},\mathbf{y})$ is a derivative of the fundamental solution
of the underlying PDE. By matching the boundary condition at
$\mathbf{x}_0 \in \partial\Omega$ with the limit of
equation~\eqref{eq:LP} as $\mathbf{x}\rightarrow \mathbf{x}_0$, a BIE
for the density function is formed. Numerical methods to solve BIEs have
several advantages over their PDE-based counterpart: by construction,
the layer potential~\eqref{eq:LP} satisfies far-field conditions; only
$\partial\Omega$ needs to be discretized; carefully chosen kernels
$K(\mathbf{x},\mathbf{y})$ result in a well-conditioned linear system
that is solved with a mesh-independent number of GMRES iterations; and
high-order or spectral accuracy is attainable with appropriate
quadrature methods. In our previous work~\cite{Fu2018_SIAM} we
demonstrate that BIEs are a powerful tool to simulate two-dimensional
suspensions of amphiphilic particles. In addition to improving
two-dimensional simulations, we will extend the results to three
dimensions using a standard three-dimensional BIE formulation of the
screened Laplace equation~\cite{ying_2006} and a well-conditioned BIE
formulation of the mobility problem~\cite{manasthesis, rac-gre2016}.

\begin{wrapfigure}[21]{r}{2.5in}
%\vspace*{+15pt}
\centerline{\includegraphics[width=2.5in]{figures/integrands}}
\vspace*{-13pt}
  \caption{{\footnotesize The Fourier modes of
  $K(\mathbf{x_0},\mathbf{y})$ and $K_2(\mathbf{x_0},\mathbf{y})$ where
  $|\mathbf{x_0}| = 0.99$ and $\mathbf{y} \in \partial\Omega$, and
  $\Omega$ is the unit circle. The trapezoid rule applied to the red
  integrand has large amounts of error, but the trapezoid rule is much
  more accurate when applied to the blue integrand. Since the difference
  between the red and blue curves can be accurately computed with the
  barycentric quadrature rule, the quadrature error applied to layer
  potentials for the screen Laplace equation will be uniformly bounded
  in $\Omega$.}}
\label{fig:integrands}
\end{wrapfigure}
Significant challenges to solve BIEs include solving dense linear
systems, developing preconditioning strategies, and developing
quadrature methods for integrands that are nearly singular. These
challenges are described in \S\ref{subsec:NumericalIssues}. However,
numerical errors can still lead to unphysical contact between bodies. We
propose two algorithms to avoid contact in \S\ref{subsec:timeStepping}.
Then, proposed work for well-conditioned BIE formulations of fluctuating
hydrodynamics are described in \S\ref{subsec:fluctuating}.

@@

Discretizations of carefully chosen BIEs can be solved with a
mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}. Therefore, the required CPU
time is proportional to the cost of a matrix-vector multiplication that
can be done in optimal or near-optimal time with the fast multipole
method (FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3,
fmm4, fmm6, fmm7, fmm8}. PI BQ is experienced with applying FMMs to
BIE formulations of the Stokes equation~\cite{qua-bir2014,
bys-sha-qua2020} and the screened Laplace equation~\cite{kro-qua2011,
qua2011}. Alternatively, fast direct solvers can be used to avoid
iteration~\cite{fds1, fds2, fds3, fds4, fds5, fds6, fds7, fds8,
ho2016cpam2, ho2016cpam1, minden2016, minden2017siammms}, but direct
solvers often require a large amount of computational overhead, and this
makes them less practical for problems with moving geometries.
Alternatively, techniques in fast direct solvers can be used to develop
efficient preconditioning strategies such as the inverse Fast Multipole
Method (IFMM)~\cite{cou-pou-dar2017}. PI BQ used the IFMM to
precondition Stokes equations in a porous media~\cite{qua-cou-dar2018},
and a suite of other preconditioners including sparse approximate
inverses~\cite{che2000} and multigrid~\cite{hem-sch1981, sch1982} will
be investigated.

Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and this results in nearly-singular integrands. Quadrature
methods for these integrands has received a lot of attention in two and
three dimensions~\cite{alpert, kapur, sidi, duffy, bruno1, bruno2,
davis_1984, graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003,
khayat_2005, schwab_1992, ying_2006, beale1, beale2, goodman_1990,
haroldson_1998, lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3,
helsing_2008a, helsing_integral_2009, helsing_tutorial_2012,
klockner2013jcp, qbx2, wala2019jcp, af2018sisc, siegel2018jcp,
rachh2017jcp, ding2019arxiv, bar2014}. The trapezoid rule is the
workhorse for two-dimensional BIEs since it achieves spectral accuracy
when integrands are not nearly-singular~\cite{tre-wei2014}. To address
nearly-singular integrands of two-dimensional BIEs, we will use a {\em
barycentric quadrature rule} that only requires a slight modification of
the trapezoid rule~\cite{ioa-pap-per1991}. In its current form, this
method requires the layer potential to satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}, but the PIs will extend
the method to layer potentials for the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as
$K(\mathbf{x},\mathbf{y}) = K_1(\mathbf{x},\mathbf{y}) +
K_2(\mathbf{x},\mathbf{y})$, where $K_1(\mathbf{x},\mathbf{y}) =
-\log|\mathbf{x} - \mathbf{y}|$ and $K_2(\mathbf{x},\mathbf{y}) =
K(\mathbf{x},\mathbf{y}) + \log|\mathbf{x} - \mathbf{y}|$. Using this
decomposition, layer potentials involving $K_1(\mathbf{x},\mathbf{y})$
can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and layer potentials involving
$K_2(\mathbf{x},\mathbf{y})$ can be accurately computed with the
trapezoid rule since the kernel is bounded for all $\mathbf{x}$ and
$\mathbf{y}$. The proposed method is demonstrated in
Figure~\ref{fig:integrands} where the amplitude of the Fourier modes of
$K(\mathbf{x_0},\mathbf{y})$ and $K_2(\mathbf{x_0},\mathbf{y})$ are
plotted. The quadrature of the trapezoid rule is the sum of the Fourier
amplitudes not captured at the resolution.
